---
title: "Chapter 1: Basic R Operation"
format: html
---

## Windows file path in R

Every time you start an R session, you will be directed to a default working directory depends on how you starting the session.

### *The Working Directory*

In order to know the path to your working directory, type

```{r}
getwd()
```

### *Set Working Directory*

You can set the working directory as below:

```{r}
#| eval: false
# Code not evaluated
setwd("D:/GitHub/R4Dummies/docs")
```

Since R uses `\` as an *escape character*, you can type the following code instead.

```{r}
#| eval: false
# Code not evaluated
setwd("D:\\GitHub\\R4Dummies\\docs")
```

### *Objects in the environment by `ls()`*

In order to see the objects in the environment, you can type the following. As no object created yet, the output is `character(0)`.

```{r}
ls()
```

## Some basic operation

We can create a *vector* by the function `c()`. The following line of code can be read as "x gets the vector 5,10,15,20,25,30,35,40." And we can type **x** to get what **x** is.

```{r}
x <- c(5,10,15,20,25,30,35,40)
x
```

Now, we have created an object **x**. We can check this by typing

```{r}
ls()
```

We can work with **x** with some basic function like the following:

```{r}
sum(x)
mean(x)
var(x)
```

## R Functions

The examples above like `c()`, `sum()`, and etc. are *functions* built into R. Insides the parentheses are *arguments*. Some functions take no arguments and some functions take more than one argument. `seq()` is an example of multiargument function. We use it to build **y**, which is the same as **x**.

```{r}
y <- seq(5,40,5)
y
```

The way to lick the arguments above is to list the arguments in the order that they appear in the function's definition. R calls this *positional mapping*. For `seq()` above, the first argument is the number to star the sequence *from* (5). The second argument is the number that ends the sequence - the number the sequence goes *to* (40). The third argument is the increment of the sequence - the amount the sequence increases *by* (5). If you *name* the arguments, it doesn't matter how you order them:

```{r}
z <- seq(to=40,by=5,from=5)
z
```

## User-Defined Functions

The form of an R function is

```{r}
#| eval: false
myfunction <- function(argument1, argument2, ...){
  statements
  return(object)
}
```

Let's create a function to calculate the length of the hypotenuse of an right angle triangle.

```{r}
hypotenuse <- function(a,b){
  hyp <- sqrt(a^2+b^2)
  return(hyp)
}
```

To use the function, type

```{r}
hypotenuse(3,4)
```

## R Structures

### *Vectors*

The *vector* is the fundamental structure in R. It's an array of elements of the same type. The data elements in a vector are called *components*.

**x**, **y**, and **z** above are samples of vector. We can create a *character vector* as follows:

```{r}
beatles <- c("john","paul","george","ringo")
beatles
```

We can create a *logical vector* as follows:

```{r}
w <- c(T,F,F,T,T,F)
w
```

To refer to a specific component of a vector, follow the vector name with a bracketed number:

```{r}
beatles[2]
beatles[2:3]    # for consecutive components
beatles[(2,4)]  # for non-consecutive components
```
